---
title: "Competition_Forecast_RMD"
author: "Nannaphat Sirison, Lynn Wu" 
output: html_document
date: "2023-04-03"
---

##  INTRODUCTION

```{r setup, include=FALSE}
#Installing required packages
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(kableExtra)
library(readxl)
library(zoo)
library(openxlsx)
library(writexl)
```

## EXPLORING THE DATA

```{r, include= FALSE}

#Loading in the data set

df<-read_excel("load.xlsx")
```

```{r, include = FALSE}

#DATA WRANGLING

#Taking the average across hours of the day

df$hourly_mean <- rowMeans(df[,3:26])

#Selecting only necessary columns: data and hourly mean

df <- df[, c("date","hourly_mean")]

#Making date column a date format

df$date <- ymd(df$date)

```

```{r, include = FALSE}
#Drop NAs
df1 <- df %>% drop_na()

#Converting to time series object (2005.1.1-2010.12.31)

ts <- msts(df1[,2], seasonal.periods = c(7,365.25),start=c(2005,01,01),end=c(2010,12,31))

```

```{r, echo= FALSE}

#Initial plot of time series
ts_plot<-ggplot(df1,aes(x=date, y=hourly_mean))+
  geom_line()
plot(ts_plot)

#Summary of time series
ts_summary<-summary(ts)
ts_summary
```

## DATA ANALYSIS

```{r}

#Original Series: ACF and PACF Plots
par(mfrow=c(1,2))
ACF_Plot <- Acf(ts, lag = 40, plot = TRUE,main="ACF of Hourly Mean")
PACF_Plot <- Pacf(ts, lag = 40,plot = TRUE,main="PACF of Hourly Mean")

```
>Plots not only show strong seasonality but also non-stationarity due to the slow decay.

```{r}
#Original Series: Decomposition
ts %>% mstl() %>%
  autoplot()
```
>The trend has a narrow range compared to seasonal components. Both weekly and yearly pattern have strong scales and trend components in daily data.

##FORECASTING DAILY ACTIVE POWER

```{r message=FALSE, warning=FALSE}
#create a subset for training purpose (2005-2009)
n_for = 365
ts_train <- subset(ts,end = length(ts)-n_for)

#create a subset for testing purpose (2009-2010)
ts_test <- subset(ts, start = length(ts)-n_for)

autoplot(ts_train)
autoplot(ts_test)
```

#Model 1: STL + ETS 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
#Fit and forecast STL + ETS model to data
ETS_fit <-  stlf(ts_train,h=365)
```

```{r}
#Plot model + observed data
autoplot(ts) +
  autolayer(ETS_fit, series="STL + ETS",PI=FALSE) +
  ylab("Hourly Load")
```

Discussion: we see that that it follows the seasonality well (as observed by the coincidental oscillation of the red and black series); however, the magnitude of the forecast does not fit the data we observe well, where observed data has a greater magnitude than the forecasted data. 

```{r, include = FALSE}
#WRANGLING FOR KAGGLE SUBMISSION

#Generate 2011 output for Kaggle competition
ETS_fit_kaggle <-  stlf(ts,h=59)
#Save ETS_fit_kaggle as a dataframe
ETS_fit_kaggle_df <- as.data.frame(ETS_fit_kaggle$mean)

#Read in submission template
submission_template<-read_excel("submission_template.xlsx")
#Merge df
submission_template_ETS <- cbind(submission_template,ETS_fit_kaggle_df)
#Remove extra column
submission_template_ETS <- submission_template_ETS[, c("date","x")]
#Rename column from "x" to "load"
colnames(submission_template_ETS) <- c("date","load")

#Save out dataframe as excel file to submit to Kaggle 
write.xlsx(submission_template_ETS, file = "Sirison_Wu_1.xlsx")

```

#Model 2: ARIMA + FOURIER terms

```{r ARIMA, echo=TRUE, message=FALSE, warning=FALSE}

ARIMA_Four_fit <- auto.arima(ts_train, 
                             seasonal=FALSE, 
                             lambda=0,
                             xreg=fourier(ts_train, 
                                          K=c(2,48))
                             )

#Forecast with ARIMA fit
ARIMA_Four_for <- forecast(ARIMA_Four_fit,
                           xreg=fourier(ts_train,
                                        K=c(2,48),
                                        h=365),
                           h=365
                           ) 

#Plot foresting results
autoplot(ARIMA_Four_for) + ylab("Hourly Load")

#Plot model + observed data
autoplot(ts) +
  autolayer(ARIMA_Four_for, series="ARIMA_FOURIER",PI=FALSE) +
  ylab("Hourly Load")
```

```{r, include = FALSE}
#WRANGLING FOR KAGGLE SUBMISSION

#Generate 2011 output for Kaggle competition
ARIMA_FOURIER_fit_kaggle <-  forecast(ARIMA_Four_for,h=59)
#Save ETS_fit_kaggle as a dataframe
ARIMA_FOURIER_fit_kaggle_df <- as.data.frame(ARIMA_FOURIER_fit_kaggle$mean)

#Merge df
submission_template_ARIMA <- cbind(submission_template,ARIMA_FOURIER_fit_kaggle_df)
#Remove extra column
submission_template_ARIMA <- submission_template_ARIMA[, c("date","x")]
#Rename column from "x" to "load"
colnames(submission_template_ARIMA) <- c("date","load_Model2")

#Merge model1 and model2
combine_model_final<- merge(submission_template_ETS,submission_template_ARIMA,by="date")
colnames(combine_model_final) <- c("date","load_Model1","load_Model2")

#Save out dataframe as excel file to submit to Kaggle 
write.xlsx(combine_model_final, file = "Sirison_Wu_2.xlsx")

```


#Model 3: TBATS

```{r, echo=TRUE, message=FALSE, warning=FALSE}
TBATS_fit <- tbats(ts_train)
TBATS_for <- forecast(TBATS_fit, h=365)
```

```{r}
#Plot TBATS forecast vs observed data 
autoplot(ts) +
  autolayer(TBATS_for, series="TBATS",PI=FALSE)+
  ylab("Hourly Load") 
```

```{r, include = FALSE}
#WRANGLING FOR KAGGLE SUBMISSION
#Generate 2011 output for Kaggle competition
TBATS_fit_kaggle <- forecast(TBATS_fit, h=59)
#Save TBATS_fit_kaggle as a dataframe
TBATS_fit_kaggle_df <- as.data.frame(TBATS_fit_kaggle$mean)

#Read in submission template
submission_template<-read_excel("submission_template.xlsx")
#Merge df
submission_template_TBATS <- cbind(submission_template,TBATS_fit_kaggle_df)
#Remove extra column
submission_template_TBATS <- submission_template_TBATS[, c("date","x")]
#Rename column from "x" to "load"
colnames(submission_template_TBATS) <- c("date","load")

#Save out dataframe as excel file to submit to Kaggle 
write.xlsx(submission_template_TBATS, file = "Sirison_Wu_3.xlsx")
```


##EXAMINING ACCURACY OF FORECASTS 

#Model 1: STL + ETS Accuracy 
```{r, include = FALSE}
ETS_scores <- accuracy(ETS_fit$mean,ts_test)  
```

#Model 2: Arima + Fourier Accuracy 
```{r, include = FALSE}
ARIMA_scores <- accuracy(ARIMA_Four_for$mean,ts_test)
```

#Model 3: TBATS Accuracy 
```{r, include = FALSE}
TBATS_scores <- accuracy(TBATS_for$mean,ts_test)
```

#Model 4: XXX Accuracy 
```{r}

```


#Accuracy Comparison
```{r}
#Graph of all forecast models 
autoplot(ts) +
  autolayer(ETS_fit, series="STL + ETS",PI=FALSE) +
  autolayer(ARIMA_Four_for, series="ARIMA_FOURIER",PI=FALSE) +
  autolayer(TBATS_for, series="TBATS",PI=FALSE)+
  ylab("Hourly Load") 
```



```{r}
#Table of accuracy statistics
```


##CONCLUSION




